---
title: "Modeling Changes in Bird Occurrence Across Residency Groups in Durham County (2015-2025)"
output:
  html_document: default
  pdf_document: default
author: Jonathan Gilman, Nicole Gutkowski, Shaochong Xue
date: "2026-04-24"
---

```{r, include=FALSE}

library(forecast)
library(Kendall)
library(tseries)
library(tidyverse)
library(readxl)
library(dplyr)
library(cowplot)
library(lubridate)
library(ggplot2)
library(smooth)
library(zoo)
library(kableExtra)
library(tidyr)
library(RColorBrewer)
library(knitr)
library(patchwork)

```


# Introduction, Motivation, Relevance, Objectives

Anthropogenic environmental change has significantly altered bird diversity and population dynamics, with particularly strong declines observed among migratory species. Long-term studies estimate that North American bird populations have declined by nearly 60% since 1970, underscoring the importance of understanding species-specific trends over time.

These impacts are not uniform across species, as different residency groups (resident, summer migrant, winter migrant) respond differently to environmental pressures. Seasonal variation in occurrence may differ substantially between these groups, making time series approaches especially valuable for capturing both long-term trends and periodic fluctuations.

This study uses Durham County, North Carolina as a case study to analyze long-term trends in bird occurrence across residency types using time series modeling. Three indicator species were selected to represent different residency strategies: Canada Goose (resident), Ruby-throated Hummingbird (summer migrant), and Yellow-rumped Warbler (winter migrant). These species were chosen based on their ecological relevance and the availability of consistent observational data.


# Dataset information

The data for this project are sourced from eBird, a global citizen science database from the Cornell Lab or Ornithology. While this study only analyzes data for three bird species in Durham, NC, the eBird dataset contains data for over 2000 species globally, making it comprehensive resource for ornithology efforts all over the world. 

For the intial download of data from eBird, the data request was filtered to our indicator species in Durham, NC. From this, we recieved a dataset with 54246 recorded observations for our three species between February 20, 2002 to January 31, 2026. The dataset included information on species identification, observation details (counts, behavior, breeding), spatial location, time, observer metadata, survey effort, and data quality/validation information.pr


# Data Cleaning (General)

Prior to species-specific analysis, several general cleaning and preparation steps were applied to the full dataset. The raw eBird data was subset to retain only three variables of interest: species identification, observation counts, and date. Although the dataset contains records dating back to 2002, analysis was restricted to January 2015 through January 2026. Data prior to 2015 was highly inconsistent, which we attributed to irregular observer effort rather than true absences of birds, making it unsuitable for reliable time series modeling.
Any dates with no recorded observations were filled with zeros, reflecting the assumption that the species was not detected on those days. Data were then aggregated to the monthly level by calculating the mean daily observation count within each month, resulting in a continuous monthly time series for each species with no gaps. Finally, each aggregated dataframe was formatted as a time series object to facilitate downstream modeling and forecasting.
Species-specific outlier removal and additional cleaning steps are described within each species' analysis section below.

# Analysis (Methods and Models)

### Canada Goose Models and Results

For the time series analysis of Canada Goose observation data, I chose to use the monthly data for training and forecasting models. Prior to any modeling, general data exploration and further cleaning were performed. 

After intitial plotting of the time series, outlier removal was performed using tsclean() (Figure X).

```{r, include = FALSE, warning= FALSE}

goose_monthly <- read_csv("./Data/Cleaned_bySpecies/CanGoose_Monthly_withObserver.csv")

goose_monthly <- goose_monthly %>% 
  mutate(month = month(ymd(date))) %>% 
  filter(date >= as.Date("2015-01-01") &
         date <= as.Date("2025-01-01"))


```


```{r, echo = FALSE, warning = FALSE, message= FALSE, fig.width=12, fig.height=5, fig.cap= "Initial time series plot and cleaning of the monthly Canada Goose Observations Data"}

ts_goose_monthly <- ts(goose_monthly$mean_obs, start = c(2015, 1), frequency = 12)

ts_plot_goose <- autoplot(ts_goose_monthly)+
  xlab("Time")+
  ylab("Average Observations (Birds)")+
  ggtitle("Monthly Canada Goose Data")+
  theme_bw(base_size= 13)+
  xlim(2015, 2025)+
  scale_x_continuous(breaks = seq(2015, 2025, by = 1))

ts_goose_cleaned <- tsclean(ts_goose_monthly)

goose_train <- ts_goose_cleaned[1:109]
ts_goose_train <- ts(goose_train, start = c(2015, 1), frequency = 12)

goose_test <- ts_goose_cleaned[110:121]
ts_goose_test <- ts(goose_test, start = c(2024, 2), frequency = 12)


ts_goose_plot_cleaned <- autoplot(ts_goose_monthly, series = "Original Data") +
  autolayer(ts_goose_cleaned, series = "Cleaned Data") +
  xlab("Time") +
  ylab("Average Observations (Birds)") +
  ggtitle("Outlier Removal: Monthly Canada Goose Data") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = c(0.02, 0.98),  
    legend.justification = c(0, 1),    
    legend.title = element_blank())

plot_grid(ts_plot_goose, ts_goose_plot_cleaned, ncol = 2)

```

After removing outliers from the data, summary statistics were also assessed (Table X). The wide range in monthly observations and large standard deviation relative to the mean indicated that while a resident species, observations may vary over the course of the year. 



```{r, echo=FALSE, warning=FALSE}

goose_clean_df <- data.frame(
  date = as.Date(time(ts_goose_cleaned)),
  mean_obs = as.numeric(ts_goose_cleaned)
)

summary_table <- goose_clean_df %>%
  summarise(
    Mean = mean(mean_obs, na.rm = TRUE),
    Median = median(mean_obs, na.rm = TRUE),
    SD = sd(mean_obs, na.rm = TRUE),
    Min = min(mean_obs, na.rm = TRUE),
    Max = max(mean_obs, na.rm = TRUE),
    `1st Quartile` = quantile(mean_obs, 0.25, na.rm = TRUE),
    `3rd Quartile` = quantile(mean_obs, 0.75, na.rm = TRUE)
  )

kbl(summary_table,
    caption = "Summary Statistics for Cleaned Monthly Goose Observations",
    digits = 2,
    booktabs = TRUE) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  )
```

To begin investigating potential seasonality or variations in observations throughout the year, the distribution of monthly observations were plotted and the series was decomposed into seasonal, trend, and residual components (Figure X).


```{r, echo = FALSE, warning = FALSE,  fig.height=4, fig.width=14, fig.cap = "Distribution of Monthly Observations of Canada Geese"}
month_colors <- c(
  "1" = "#4575b4",  # Jan - cold blue
  "2" = "#74add1",
  "3" = "#abd9e9",
  "4" = "#e0f3f8",
  "5" = "#fee090",
  "6" = "#fdae61",
  "7" = "#f46d43",
  "8" = "#f46d43",  # peak summer - hot red
  "9" = "#fdae61",
  "10" = "#fee090",
  "11" = "#abd9e9",
  "12" = "#4575b4"  # Dec - cold again
)

goose_monthly$month <- factor(goose_monthly$month, levels = 1:12)

goose_monthly_dist <- ggplot(goose_monthly, aes(x = month, y = mean_obs, fill = month))+
  geom_boxplot()+
  ylim(0, 200)+
  theme_bw(base_size= 13)+
  scale_fill_manual(values = month_colors)+
  ggtitle("Monthly Canada Goose Observations in Durham (2015-2025)")+
  ylab("Average Observations (Birds)")+
  xlab("Month")+
  theme(legend.position = "none")+ 
  scale_x_discrete(labels = month.abb)

ts_goose_decomp = decompose(ts_goose_cleaned, type = "additive")

goose_decomp_plot <- autoplot(ts_goose_decomp)+
  ggtitle("Decomposition of Cleaned Canada Goose Data")+
  theme_bw(base_size= 13)

plot_grid(goose_monthly_dist, goose_decomp_plot, ncol = 2)

``` 
From the plot of distributions of monthly observations, the Canada Goose appears to have a slight seasonal pattern. However, consistency in observations year-round support the Canada Goose's resident species status. 

The decomposition of the cleaned series also shows a strong seasonal pattern and trend. However the trend does not appear to have a consistent pattern over the course of the time series.  

#### ACF and PACF

The ACF and PACF were then plotted and assessed for initial modeling considerations (Figure X).

```{r, echo = FALSE, warning= FALSE, fig.cap= "ACF and PACF plots for the cleaned monthly Canada Goose time series"}
a <- autoplot(Acf(ts_goose_cleaned, lag.max = 50, plot = FALSE))+
  ggtitle(" ")+
  theme_bw(base_size= 13)
b <- autoplot(Pacf(ts_goose_cleaned, lag.max = 50, plot = FALSE))+
  ggtitle(" ")+
  theme_bw(base_size= 13)

plot_row <- plot_grid(a, b, ncol = 2)

# now add the title
title <- ggdraw() + 
  draw_label(
    "Cleaned Canada Goose Data: ACF and PACF",
    x = 0,
    hjust = 0,
    vjust = 1
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 0)
  )
plot_grid(
  title, plot_row,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.09, 1)
)


```

The ACF plot for the series shows a strong seasonality at lag 12, suggesting that models will either need to account for seasonality, or seasonality should be removed prior to modeling. The gradual decacy of the series also suggests stationarity, and that the series may require differencing. 

The PACF plot for the series has a clear cut off with only 1 significant lag. This behavior suggests the series has an autoregressive component. 

Based on observations from the decomposition of the series and the ACF and PACF, we thought that models accounting for seasonality would be best for the series. The initial models that were chosen were the SARIMA and the Basic Structural Model.  

    
SARIMA

The SARIMA model was chosen for forecasting our series due to it's widely used and well established approach to time series analysis with seasonality. The autofit SARIMA model was calculated to be SARIMA(1,1,1)(1,0,0)[12]. The forecast of our data compared with the true observations for 2025 is shown in Figure X. 


```{r, echo = FALSE, warning= FALSE, message = FALSE}
auto_sarima111_100_fit <- auto.arima(ts_goose_train)
auto_sarima111_100_for <- forecast(auto_sarima111_100_fit, h = 12)

autoplot(auto_sarima111_100_for, series = "Auto-SARIMA")+
  autolayer(ts_goose_test, series = "True Observations")+
  theme_bw(base_size= 13)+
  xlim(2015, 2025)+
  scale_x_continuous(breaks = seq(2015, 2025, by = 5))+
  xlab("Time")+
  ylab("Average Observations (Birds)")+
  ggtitle("Auto-SARIMA Forecast") +
  theme(
    legend.position = c(0.02, 0.98),  
    legend.justification = c(0, 1),    
    legend.title = element_blank())

sarima111_100_scores <- accuracy(auto_sarima111_100_for$mean,goose_test)
sarima111_100_scores


```


#### BSM model

The second model that was chosen for forecasting the Canada Goose data was the Basic Structural Model. This model was chosen due to its ability to both handle seasonality and to account for a localized trend. Because there did not appear to be a clear directional trend in the series, the BSM model was potentially a solution for modeling. The 2025 forecast of the BSM model compared with the true observations for 2025 is shown in Figure X. 

```{r, echo = FALSE, warning= FALSE, message=FALSE}

structs_nafix <- StructTS(ts_goose_train, type = "BSM",
                         fixed=c(NA,NA,NA,NA))
#Variances:
#  level    slope   seas     epsilon  
# 431.92     0.00    11.17   117.16


#this function has convergence issues
#checkresiduals(structs_nafix)

#Generating forecasts
# StructTS() does not call the forecast() internally so we need one more step
auto_ss_for <- forecast(structs_nafix,h=12)
#plot(auto_ss_for)

SS_for1_score <- accuracy(auto_ss_for$mean,goose_test)
SS_for1_score

autoplot(auto_ss_for, series = "Auto-BSM")+
  autolayer(ts_goose_test, series = "True Observations")+
  theme_bw(base_size= 13)+
  xlim(2015, 2025)+
  scale_x_continuous(breaks = seq(2015, 2025, by = 5))+
  xlab("Time")+
  ylab("Average Observations (Birds)")+
  ggtitle("Auto-BSM Forecast") +
  theme(
    legend.position = c(0.02, 0.98),  
    legend.justification = c(0, 1),    
    legend.title = element_blank())


```

#### Comparison of Models

After fitting and forecasting the SARIMA and BSM models separately, the results of the forecasting were plotted together for intitial comparison (Figure X). While both models were able to capture the general seasonality of the data, neither were able to predict the magnitude of the seasonal dip in the series. 

```{r, echo = FALSE, warning= FALSE, message=FALSE, fig.cap= "Comparison of forecasts for 2025 Canada Goose Observations"}

autoplot(ts_goose_test, series = "True Observations")+
  autolayer(auto_sarima111_100_for,PI=FALSE, series = "Auto-SARIMA")+
  autolayer(auto_ss_for,PI=FALSE, series= "Auto-BSM")+
  theme_bw(base_size= 13)+
  xlim(2024, 2025)+
  scale_x_continuous(breaks = seq(2024, 2025, by = 1))+
  xlab("Test Period (2024)")+
  ylab("Average Observations (Birds)")+
  ggtitle("Comparison Average Monthly Observation Forecasts")


```
To further compare the results of the forecasting models, error metrics were calculated for both models (Table X). These metrics included: Mean Error (ME), Root Mean Square Error (RMSE), Mean Absolute Error (MAE), Mean Percent Error (MPE), and Mean Absolute Percent Error (MAPE). The SARIMA model outperformed the BSM model in all error metrics with the exception of Mean Error (ME), making it the top performing model for the Canada Goose monthly observation data. 

```{r, echo = FALSE, warning= FALSE}

goose_scores <- as.data.frame(rbind(SS_for1_score, sarima111_100_scores))
row.names(goose_scores) <- c("Auto-BSM", "Auto-SARIMA")

#choose model with lowest RMSE
#best_model_index_goose <- which.min(goose_scores[,"RMSE"])
#cat("The best model by RMSE is:", row.names(goose_scores[best_model_index_goose,])) 


kbl(goose_scores,
    caption = "Forecasts for Monthly Canada Goose Observations",
    digits = array(2, ncol(goose_scores))) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    latex_options = c("hold_position", "striped")
  )

```



### Warbler Models and Results

The Yellow-rumped Warbler (YRWA) is a common winter migratory species in North Carolina. In this section, we analyze the temporal pattern of YRWA observations and develop time series models to forecast future trends.

We compare modeling results based on observation count and observer ratio. The observer ratio standardizes observation counts and helps control for variation in birding effort over time.

```{r, , echo = FALSE, warning = FALSE}
# Monthly Data
YRWA_monthly <- read.csv("./Data/Cleaned_bySpecies/Warbler_Monthly.csv")
YRWA_monthly_obs <- read.csv("./Data/Cleaned_bySpecies/Warbler_Monthly_withObserver.csv")

# filter observation data from 2015-01-01 to 2025-12-31
YRWA_monthly <- YRWA_monthly %>% 
  mutate(
    date = ymd(date),
    month = month(date)
  ) %>% 
  filter(date >= as.Date("2015-01-01") &
         date <= as.Date("2025-12-31"))

YRWA_monthly_obs <- YRWA_monthly_obs %>% 
  mutate(
    date = ymd(date),
    month = month(date)
  ) %>% 
  filter(date >= as.Date("2015-01-01") &
         date <= as.Date("2025-12-31"))

# convert to ts
YRWA_obs_ts <- ts(
  YRWA_monthly$mean_obs,
  start = c(2015, 1),
  frequency = 12
)
```


```{r, echo = FALSE, warning = FALSE}
# Monthly observer ratio
# total observer
bird_data <- read.csv("./Data/Raw/eBird_Species_Data.csv")

# daily total observers
daily_total_observers <- bird_data %>%
  mutate(
    OBSERVATION.DATE = as.Date(OBSERVATION.DATE),
    date = OBSERVATION.DATE
  ) %>%
  filter(
    date >= as.Date("2015-01-01") &
    date <= as.Date("2025-12-31")
  ) %>%
  group_by(date) %>%
  summarise(
    total_observers = n_distinct(OBSERVER.ID),
    .groups = "drop"
  )

# daily observer
YRWA_daily_obs <- read.csv("./Data/Cleaned_bySpecies/Warbler_Daily_withObserver.csv")

YRWA_daily_obs <- YRWA_daily_obs %>%
  mutate(
    date = ymd(date),
    month = month(date)
  ) %>% 
  filter(date >= as.Date("2015-01-01") &
         date <= as.Date("2025-12-31"))

# daily ratio
YRWA_daily_ratio <- daily_total_observers %>%
  left_join(YRWA_daily_obs, by = "date") %>%
  mutate(
    sum_observers = ifelse(is.na(sum_observers), 0, sum_observers),
    sum_obs = ifelse(is.na(sum_obs), 0, sum_obs),
    observer_ratio = ifelse(total_observers == 0, 0, sum_observers / total_observers)
  )

# monthly ratio
YRWA_monthly_ratio <- YRWA_daily_ratio %>%
  mutate(date = floor_date(date, "month")) %>% 
  group_by(date) %>%
  summarise(
    mean_obs = mean(sum_obs),
    mean_observers = mean(sum_observers),
    mean_observer_ratio = mean(observer_ratio),
    .groups = "drop"
  )

# convert to monthly ts
YRWA_ratio_ts <- ts(
  YRWA_monthly_ratio$mean_observer_ratio,
  start = c(2015, 1),
  frequency = 12
)
```

#### Summary Statistics

```{r, echo = FALSE, warning = FALSE}
obs_summary_table <- data.frame(
  Dataset = "Observations",
  Mean = mean(as.numeric(YRWA_obs_ts), na.rm = TRUE),
  Median = median(as.numeric(YRWA_obs_ts), na.rm = TRUE),
  SD = sd(as.numeric(YRWA_obs_ts), na.rm = TRUE),
  Min = min(as.numeric(YRWA_obs_ts), na.rm = TRUE),
  Max = max(as.numeric(YRWA_obs_ts), na.rm = TRUE),
  `1st Quartile` = quantile(as.numeric(YRWA_obs_ts), 0.25, na.rm = TRUE),
  `3rd Quartile` = quantile(as.numeric(YRWA_obs_ts), 0.75, na.rm = TRUE)
)

ratio_summary_table <- data.frame(
  Dataset = "Observer Ratio",
  Mean = mean(as.numeric(YRWA_ratio_ts), na.rm = TRUE),
  Median = median(as.numeric(YRWA_ratio_ts), na.rm = TRUE),
  SD = sd(as.numeric(YRWA_ratio_ts), na.rm = TRUE),
  Min = min(as.numeric(YRWA_ratio_ts), na.rm = TRUE),
  Max = max(as.numeric(YRWA_ratio_ts), na.rm = TRUE),
  `1st Quartile` = quantile(as.numeric(YRWA_ratio_ts), 0.25, na.rm = TRUE),
  `3rd Quartile` = quantile(as.numeric(YRWA_ratio_ts), 0.75, na.rm = TRUE)
)

combined_summary_table <- bind_rows(obs_summary_table, ratio_summary_table)

kbl(
  combined_summary_table,
  caption = "Summary Statistics for Monthly YRWA Observations and Observer Ratio",
  digits = 2,
  booktabs = TRUE,
  row.names = FALSE
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  )
```

Summary statistics show clear differences between observation count and observer ratio. The mean monthly observation count is 15.80, with a relatively large standard deviation (15.27), indicating high variability across months. In contrast, the observer ratio has a lower mean (0.43) and smaller variability (SD = 0.36), suggesting a more stable relative measure after accounting for observer effort.

Both variables include values close to zero, reflecting months with very low or no observations. The maximum observation count reaches 53.2, while the observer ratio peaks at 0.9, indicating periods of high detection relative to effort.

```{r, echo = FALSE, warning = FALSE}
p1 <- autoplot(YRWA_obs_ts) +
  xlab(NULL) +
  ylab("Mean Observations") +
  ggtitle("Observations") +
  theme_bw(base_size = 12)

p2 <- autoplot(YRWA_ratio_ts) +
  xlab("Time") +
  ylab("Observer Ratio") +
  ggtitle("Observer Ratio") +
  theme_bw(base_size = 12)

p1 / p2 +
  plot_annotation(
    title = "Monthly YRWA Observations and Observer Ratio (2015–2025)"
  )
```

```{r, echo = FALSE, warning = FALSE}
autoplot(YRWA_obs_ts, series = "Observations") +
  autolayer(YRWA_ratio_ts * max(YRWA_obs_ts), 
            series = "Observer Ratio (scaled)") +
  xlab("Time") +
  ylab("Value") +
  ggtitle("Comparison: Observations vs Observer Ratio") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.title = element_blank()
  )
```

The time series plots reveal strong and consistent seasonal patterns in both observation count and observer ratio. Peaks occur regularly each year, while values drop close to zero during off-season months. 

Although the observation count shows greater variability and more pronounced spikes, the observer ratio follows a very similar seasonal pattern, indicating that the temporal dynamics are not solely driven by changes in observer effort.

When comparing observation count with the scaled observer ratio, the two series closely align in terms of timing and overall pattern. However, the observation count exhibits larger fluctuations, while the observer ratio appears smoother and more consistent. This highlights the advantage of the observer ratio in reducing the influence of varying sampling effort.

```{r, echo = FALSE, warning = FALSE}
#box plot

obs_df <- YRWA_monthly %>%
  select(month, value = mean_obs) %>%
  mutate(type = "Observations")

ratio_df <- YRWA_daily_ratio %>%
  select(month, value = observer_ratio) %>%
  mutate(type = "Observer Ratio")

plot_df <- bind_rows(obs_df, ratio_df)

plot_df$month <- factor(plot_df$month, levels = 1:12)

# plot
ggplot(plot_df, aes(x = month, y = value, fill = month)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~type, scales = "free_y") +
  theme_bw(base_size = 13) +
  scale_fill_manual(values = month_colors) +
  scale_x_discrete(labels = month.abb) +
  labs(
    title = "Seasonal Patterns: Observations vs Observer Ratio",
    x = "Month",
    y = NULL
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```
The seasonal boxplots further confirm strong monthly patterns. Observations are highest during April, corresponding to peak migration periods. In contrast, values drop to near zero during summer months (June to August), when the species is largely absent from the region.

The observer ratio shows a nearly identical seasonal structure, with high values concentrated in winter months and minimal activity in summer. This consistency suggests that the observed seasonal pattern reflects true biological behavior rather than sampling bias.

Additionally, variability is highest during peak months, indicating that observation intensity can vary substantially across years during migration periods.

```{r,  echo = FALSE, warning = FALSE}
# STL
YRWA_obs_stl <- stl(YRWA_obs_ts, s.window = "periodic")
YRWA_ratio_stl <- stl(YRWA_ratio_ts, s.window = "periodic")

# plots
p_obs <- autoplot(YRWA_obs_stl) +
  ggtitle("Observations") +
  theme_bw(base_size = 11)

p_ratio <- autoplot(YRWA_ratio_stl) +
  ggtitle("Observer Ratio") +
  theme_bw(base_size = 11)

# combine
(p_obs / p_ratio) +
  plot_annotation(
    title = "STL Decomposition: YRWA Observations vs Observer Ratio"
  ) +
  plot_layout(heights = c(1, 1.2))
```


#### ACF and PACF

```{r, echo = FALSE, warning = FALSE}
# Observation ACF
a <- autoplot(
  Acf(YRWA_obs_ts, lag.max = 50, plot = FALSE)
) +
  ggtitle("") +
  theme_bw(base_size = 13)

# Observation PACF
b <- autoplot(
  Pacf(YRWA_obs_ts, lag.max = 50, plot = FALSE)
) +
  ggtitle("") +
  theme_bw(base_size = 13)

# combine row
plot_row <- plot_grid(a, b, ncol = 2)

# title
title <- ggdraw() +
  draw_label(
    "YRWA Observations: ACF and PACF",
    x = 0,
    hjust = 0,
    vjust = 1
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )

plot_grid(
  title, plot_row,
  ncol = 1,
  rel_heights = c(0.08, 1)
)
```

```{r, echo = FALSE, warning = FALSE}
# ratio ACF
c <- autoplot(
  Acf(YRWA_ratio_ts, lag.max = 50, plot = FALSE)
) +
  ggtitle("") +
  theme_bw(base_size = 13)

# ratio PACF
d <- autoplot(
  Pacf(YRWA_ratio_ts, lag.max = 50, plot = FALSE)
) +
  ggtitle("") +
  theme_bw(base_size = 13)

# combine row
plot_ratio_row <- plot_grid(c, d, ncol = 2)

# title
title <- ggdraw() +
  draw_label(
    "YRWA Observer Ratio: ACF and PACF",
    x = 0,
    hjust = 0,
    vjust = 1
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )

plot_grid(
  title, plot_ratio_row,
  ncol = 1,
  rel_heights = c(0.08, 1)
)
```

The ACF plots for both observation count and observer ratio show strong and repeated spikes at seasonal lags (e.g., 12, 24, and 36 months), indicating clear annual seasonality in the data. In both cases, autocorrelation decays slowly, suggesting persistent temporal structure rather than random variation.

The PACF plots for both series show a strong spike at lag 1, along with weaker signals at seasonal lags. This indicates the presence of short-term dependence as well as seasonal effects.

Overall, both observation count and observer ratio exhibit similar autocorrelation structures, suggesting that the seasonal pattern is robust and not driven solely by variation in observer effort. These results support the use of seasonal time series models such as SARIMA.

#### YRWA Observation Time Series Modeling and Forecasting

We fit three time series models to the monthly YRWA observation series: a seasonal ARIMA (SARIMA) model, an ARIMA model with Fourier terms, and an exponential smoothing (ETS) model. The SARIMA model captures both short-term dependence and seasonal structure, while the ARIMA + Fourier model provides a flexible way to model smooth seasonal patterns. The ETS model serves as a benchmark based on error, trend, and seasonal components.

To evaluate model performance, we split the time series into a training set (ending in December 2024) and a test set (starting in January 2025). Each model was fitted on the training data and used to generate forecasts for the test period. Model adequacy was assessed using residual diagnostics, and forecasting performance on test set.

```{r, echo = FALSE, warning = FALSE}
# fit modle on full dataset
# SARIMA
fit_obs_sarima <- auto.arima(
  YRWA_obs_ts,
  seasonal = TRUE
)

# ARIMA + Fourier
fit_obs_fourier <- auto.arima(
  YRWA_obs_ts,
  seasonal = FALSE,
  xreg = fourier(YRWA_obs_ts, K = 2)
)

# ETS
fit_obs_ets <- ets(YRWA_obs_ts)

# Store models in a list for cleaner workflow
obs_models_full <- list(
  SARIMA = fit_obs_sarima,
  `ARIMA + Fourier` = fit_obs_fourier,
  ETS = fit_obs_ets
)
```

```{r, include = FALSE, warning = FALSE}
checkresiduals(fit_obs_sarima)
checkresiduals(fit_obs_fourier)
checkresiduals(fit_obs_ets)
```

```{r, echo = FALSE, warning = FALSE}
# train test split
obs_train_ts <- window(YRWA_obs_ts, end = c(2024, 12))
obs_test_ts  <- window(YRWA_obs_ts, start = c(2025, 1))

h <- length(obs_test_ts)
```

```{r, echo = FALSE, warning = FALSE}
# fit model on training dataset
fit_obs_sarima_train <- auto.arima(
  obs_train_ts,
  seasonal = TRUE
)

fit_obs_fourier_train <- auto.arima(
  obs_train_ts,
  seasonal = FALSE,
  xreg = fourier(obs_train_ts, K = 2)
)

fit_obs_ets_train <- ets(obs_train_ts)

obs_models_train <- list(
  SARIMA = fit_obs_sarima_train,
  `ARIMA + Fourier` = fit_obs_fourier_train,
  ETS = fit_obs_ets_train
)
```

```{r, echo = FALSE, warning = FALSE}
# test set forecasting
fc_obs_sarima <- forecast(
  fit_obs_sarima_train,
  h = h
)

fc_obs_fourier <- forecast(
  fit_obs_fourier_train,
  xreg = fourier(obs_train_ts, K = 2, h = h)
)

fc_obs_ets <- forecast(
  fit_obs_ets_train,
  h = h
)

obs_forecasts <- list(
  SARIMA = fc_obs_sarima,
  `ARIMA + Fourier` = fc_obs_fourier,
  ETS = fc_obs_ets
)
```

```{r, echo = FALSE, warning = FALSE}
plot_obs_forecast <- function(fc_obj, actual_ts, title_text) {
  autoplot(fc_obj, include = 36) + 
    autolayer(actual_ts, series = "Actual", linewidth = 0.9) +
    xlab("Time") +
    ylab("Mean Observations") +
    ggtitle(title_text) +
    theme_bw(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.title = element_blank(),
      panel.grid.minor = element_blank()
    )
}

p_sarima <- plot_obs_forecast(
  fc_obs_sarima,
  obs_test_ts,
  "SARIMA"
)

p_fourier <- plot_obs_forecast(
  fc_obs_fourier,
  obs_test_ts,
  "ARIMA + Fourier"
)

p_ets <- plot_obs_forecast(
  fc_obs_ets,
  obs_test_ts,
  "ETS"
)

options(repr.plot.width = 14, repr.plot.height = 4)

p_sarima + p_fourier + p_ets +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Forecast Comparison: YRWA Observations"
  ) &
  theme(legend.position = "bottom")
```

```{r, echo = FALSE, warning = FALSE}
obs_plot_df <- data.frame(
  Time = time(obs_test_ts),
  Actual = as.numeric(obs_test_ts),
  SARIMA = as.numeric(fc_obs_sarima$mean),
  ARIMA_Fourier = as.numeric(fc_obs_fourier$mean),
  ETS = as.numeric(fc_obs_ets$mean)
) %>%
  pivot_longer(
    cols = c("SARIMA", "ARIMA_Fourier", "ETS"),
    names_to = "Model",
    values_to = "Forecast"
  )

ggplot(obs_plot_df, aes(x = Time)) +
  geom_line(aes(y = Actual), linewidth = 1.1, color = "black") +
  geom_line(aes(y = Forecast, color = Model), linewidth = 1) +
  xlab("Time") +
  ylab("Mean Observations") +
  ggtitle("Observed vs Forecasted YRWA Observations") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

The forecast results show that all three models capture the strong seasonal pattern in YRWA observations, with similar timing of peaks and troughs. However, their accuracy differs in detail. The SARIMA model aligns most closely with the observed values across the test period, particularly in capturing both the magnitude and timing of seasonal peaks. In comparison, the ARIMA + Fourier model tends to produce more extreme fluctuations and shows larger deviations from the observed series, while the ETS model provides a smoother fit but slightly underestimates peak values. Overall, SARIMA provides the most balanced and accurate representation of the observed data.

```{r, echo = FALSE, warning = FALSE}
obs_sarima_acc  <- accuracy(fc_obs_sarima,  obs_test_ts)[2, c("RMSE", "MAE", "MAPE", "MASE")]
obs_fourier_acc <- accuracy(fc_obs_fourier, obs_test_ts)[2, c("RMSE", "MAE", "MAPE", "MASE")]
obs_ets_acc     <- accuracy(fc_obs_ets,     obs_test_ts)[2, c("RMSE", "MAE", "MAPE", "MASE")]

obs_acc_df <- data.frame(
  Model = c("SARIMA", "ARIMA + Fourier", "ETS"),
  RMSE  = c(as.numeric(obs_sarima_acc["RMSE"]),
            as.numeric(obs_fourier_acc["RMSE"]),
            as.numeric(obs_ets_acc["RMSE"])),
  MAE   = c(as.numeric(obs_sarima_acc["MAE"]),
            as.numeric(obs_fourier_acc["MAE"]),
            as.numeric(obs_ets_acc["MAE"])),
  MAPE  = c(as.numeric(obs_sarima_acc["MAPE"]),
            as.numeric(obs_fourier_acc["MAPE"]),
            as.numeric(obs_ets_acc["MAPE"])),
  MASE  = c(as.numeric(obs_sarima_acc["MASE"]),
            as.numeric(obs_fourier_acc["MASE"]),
            as.numeric(obs_ets_acc["MASE"]))
)

obs_acc_df <- obs_acc_df %>%
  mutate(across(c(RMSE, MAE, MAPE, MASE), ~ round(.x, 3)))

knitr::kable(
  obs_acc_df,
  caption = "Forecast Accuracy for YRWA Observation Models",
  digits = 3
) %>%
  kableExtra::kable_styling(full_width = FALSE)
```

```{r, include = FALSE, warning = FALSE}
best_obs_model <- obs_acc_df$Model[which.min(obs_acc_df$RMSE)]
best_obs_model
```

The forecast accuracy results show that the SARIMA model performs best among the three models, with the lowest RMSE (12.604), MAE (9.807), and MASE (1.550). In comparison, the ARIMA + Fourier and ETS models both have higher error values, indicating weaker out-of-sample performance. Although MAPE is infinite for all models due to zero values in the test set, the relative ranking based on other metrics clearly favors the SARIMA model.

```{r, echo = FALSE, warning = FALSE}
fit_obs_final <- auto.arima(
  YRWA_obs_ts,
  seasonal = TRUE
)

summary(fit_obs_final)
checkresiduals(fit_obs_final)

fc_obs_final <- forecast(
  fit_obs_final,
  h = 12
)

autoplot(fc_obs_final) +
  xlab("Time") +
  ylab("Mean Observations") +
  ggtitle("Final 12-Month Forecast: YRWA Monthly Observations") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.minor = element_blank()
  )
```
The selected SARIMA model (ARIMA(1,0,0)(2,1,0)[12]) captures both short-term dependence and strong seasonal structure in the data. Residual diagnostics indicate that the model adequately accounts for temporal patterns, as the residuals fluctuate around zero and show no clear systematic structure. The ACF of residuals does not display significant autocorrelation, and the Ljung–Box test (p-value = 0.8967) suggests that the residuals are not significantly different from white noise. This indicates that the model provides a good fit to the observed time series.

The final 12-month forecast based on the SARIMA model shows a continuation of the strong seasonal pattern observed in the historical data. Predicted values rise and fall in a regular annual cycle, reflecting the migration-driven dynamics of YRWA observations. The prediction intervals widen over time, indicating increasing uncertainty in longer-term forecasts, but the overall seasonal structure remains stable. This suggests that future observations are expected to follow similar seasonal patterns as in previous years.

#### YRWA Observer Ratio Time Series Modeling and Forecasting

We applied a similar modeling framework to the YRWA observer ratio time series as used for the observation data. Specifically, we fitted SARIMA, ARIMA with Fourier terms, and ETS models, and evaluated their performance using a train–test split and forecasting.

```{r, echo=FALSE, warning=FALSE}
# fit model on full dataset
# SARIMA
fit_ratio_sarima <- auto.arima(
  YRWA_ratio_ts,
  seasonal = TRUE
)

# ARIMA + Fourier
fit_ratio_fourier <- auto.arima(
  YRWA_ratio_ts,
  seasonal = FALSE,
  xreg = fourier(YRWA_ratio_ts, K = 2)
)

# ETS
fit_ratio_ets <- ets(YRWA_ratio_ts)

# Store models in a list for cleaner workflow
ratio_models_full <- list(
  SARIMA = fit_ratio_sarima,
  `ARIMA + Fourier` = fit_ratio_fourier,
  ETS = fit_ratio_ets
)
```

```{r, include=FALSE, warning=FALSE}
checkresiduals(fit_ratio_sarima)
checkresiduals(fit_ratio_fourier)
checkresiduals(fit_ratio_ets)
```

```{r, echo=FALSE, warning=FALSE}
# train test split
ratio_train_ts <- window(YRWA_ratio_ts, end = c(2024, 12))
ratio_test_ts  <- window(YRWA_ratio_ts, start = c(2025, 1))

h_ratio <- length(ratio_test_ts)
```

```{r, echo=FALSE, warning=FALSE}
# fit model on training dataset
fit_ratio_sarima_train <- auto.arima(
  ratio_train_ts,
  seasonal = TRUE
)

fit_ratio_fourier_train <- auto.arima(
  ratio_train_ts,
  seasonal = FALSE,
  xreg = fourier(ratio_train_ts, K = 2)
)

fit_ratio_ets_train <- ets(ratio_train_ts)

ratio_models_train <- list(
  SARIMA = fit_ratio_sarima_train,
  `ARIMA + Fourier` = fit_ratio_fourier_train,
  ETS = fit_ratio_ets_train
)
```

```{r, echo=FALSE, warning=FALSE}
# test set forecasting
fc_ratio_sarima <- forecast(
  fit_ratio_sarima_train,
  h = h_ratio
)

fc_ratio_fourier <- forecast(
  fit_ratio_fourier_train,
  xreg = fourier(ratio_train_ts, K = 2, h = h_ratio)
)

fc_ratio_ets <- forecast(
  fit_ratio_ets_train,
  h = h_ratio
)

ratio_forecasts <- list(
  SARIMA = fc_ratio_sarima,
  `ARIMA + Fourier` = fc_ratio_fourier,
  ETS = fc_ratio_ets
)
```

```{r, echo=FALSE, warning=FALSE}
plot_ratio_forecast <- function(fc_obj, actual_ts, title_text) {
  autoplot(fc_obj, include = 36) +
    autolayer(actual_ts, series = "Actual", linewidth = 0.9) +
    xlab("Time") +
    ylab("Observer Ratio") +
    ggtitle(title_text) +
    theme_bw(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.title = element_blank(),
      panel.grid.minor = element_blank()
    )
}

p_ratio_sarima <- plot_ratio_forecast(
  fc_ratio_sarima,
  ratio_test_ts,
  "SARIMA"
)

p_ratio_fourier <- plot_ratio_forecast(
  fc_ratio_fourier,
  ratio_test_ts,
  "ARIMA + Fourier"
)

p_ratio_ets <- plot_ratio_forecast(
  fc_ratio_ets,
  ratio_test_ts,
  "ETS"
)

options(repr.plot.width = 14, repr.plot.height = 4)

p_ratio_sarima + p_ratio_fourier + p_ratio_ets +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Forecast Comparison: YRWA Observer Ratio"
  ) &
  theme(legend.position = "bottom")
```

```{r, echo=FALSE, warning=FALSE}
ratio_plot_df <- data.frame(
  Time = time(ratio_test_ts),
  Actual = as.numeric(ratio_test_ts),
  SARIMA = as.numeric(fc_ratio_sarima$mean),
  ARIMA_Fourier = as.numeric(fc_ratio_fourier$mean),
  ETS = as.numeric(fc_ratio_ets$mean)
) %>%
  pivot_longer(
    cols = c("SARIMA", "ARIMA_Fourier", "ETS"),
    names_to = "Model",
    values_to = "Forecast"
  )

ggplot(ratio_plot_df, aes(x = Time)) +
  geom_line(aes(y = Actual), linewidth = 1.1, color = "black") +
  geom_line(aes(y = Forecast, color = Model), linewidth = 1) +
  xlab("Time") +
  ylab("Observer Ratio") +
  ggtitle("Observed vs Forecasted YRWA Observer Ratio") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

The forecast results for the observer ratio show that all three models capture the strong seasonal pattern, with similar timing of peaks and troughs. Compared to the observation count, the forecasts appear more stable and consistent across models, reflecting the effect of standardizing for observer effort. Among the models, SARIMA and ETS align more closely with the observed values, while the ARIMA + Fourier model shows slightly larger deviations, particularly during low-value periods. Overall, the differences between models are smaller than in the observation-based results, suggesting that the observer ratio provides a smoother and more stable signal for forecasting.

```{r, echo=FALSE, warning=FALSE}
ratio_sarima_acc  <- accuracy(fc_ratio_sarima,  ratio_test_ts)[2, c("RMSE", "MAE", "MAPE", "MASE")]
ratio_fourier_acc <- accuracy(fc_ratio_fourier, ratio_test_ts)[2, c("RMSE", "MAE", "MAPE", "MASE")]
ratio_ets_acc     <- accuracy(fc_ratio_ets,     ratio_test_ts)[2, c("RMSE", "MAE", "MAPE", "MASE")]

ratio_acc_df <- data.frame(
  Model = c("SARIMA", "ARIMA + Fourier", "ETS"),
  RMSE  = c(as.numeric(ratio_sarima_acc["RMSE"]),
            as.numeric(ratio_fourier_acc["RMSE"]),
            as.numeric(ratio_ets_acc["RMSE"])),
  MAE   = c(as.numeric(ratio_sarima_acc["MAE"]),
            as.numeric(ratio_fourier_acc["MAE"]),
            as.numeric(ratio_ets_acc["MAE"])),
  MAPE  = c(as.numeric(ratio_sarima_acc["MAPE"]),
            as.numeric(ratio_fourier_acc["MAPE"]),
            as.numeric(ratio_ets_acc["MAPE"])),
  MASE  = c(as.numeric(ratio_sarima_acc["MASE"]),
            as.numeric(ratio_fourier_acc["MASE"]),
            as.numeric(ratio_ets_acc["MASE"]))
)

ratio_acc_df <- ratio_acc_df %>%
  mutate(across(c(RMSE, MAE, MAPE, MASE), ~ round(.x, 3)))

knitr::kable(
  ratio_acc_df,
  caption = "Forecast Accuracy for YRWA Observer Ratio Models",
  digits = 3
) %>%
  kableExtra::kable_styling(full_width = FALSE)
```

```{r, echo=FALSE, warning=FALSE}
best_ratio_model <- ratio_acc_df$Model[which.min(ratio_acc_df$RMSE)]

fit_ratio_final <- auto.arima(
  YRWA_ratio_ts,
  seasonal = TRUE
)

summary(fit_ratio_final)
checkresiduals(fit_ratio_final)

fc_ratio_final <- forecast(
  fit_ratio_final,
  h = 12
)

autoplot(fc_ratio_final) +
  xlab("Time") +
  ylab("Observer Ratio") +
  ggtitle("Final 12-Month Forecast: YRWA Observer Ratio") +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.minor = element_blank()
  )
```

The forecast accuracy results show that the SARIMA model performs best for the observer ratio series, with the lowest RMSE, MAE, and MASE. This result is consistent with the findings for the observation data. The ETS model performs slightly worse but remains comparable, while the ARIMA + Fourier model shows larger errors. 

The selected SARIMA model captures the strong seasonal pattern in the observer ratio series, and residual diagnostics indicate a good fit with no significant remaining autocorrelation. The final 12-month forecast shows a clear continuation of the annual cycle, with smoother and more stable patterns than those observed in the raw observation data. Although uncertainty increases over time, the seasonal structure remains consistent, indicating that future trends are likely to follow similar dynamics.


### Ruby-throated Hummingbird Models and Results

For the Ruby-throated Hummingbird analysis, I used monthly data from 2015-2025. The Ruby-throated Hummingbird is a summer migrant that breeds in Durham and migrates south for winter. Prior to modeling, I performed data exploration and cleaning.

```{r, include=FALSE, warning=FALSE}
# Load data
humm_monthly <- read_csv("./Data/Cleaned_bySpecies/Hummingbird_Monthly_withObserver.csv")

humm_monthly <- humm_monthly %>%
  mutate(date = ymd(date),
         month = month(date)) %>%
  filter(date >= as.Date("2015-01-01") & 
         date <= as.Date("2025-01-01"))

# Train-test split
humm_train <- humm_monthly %>% filter(date <= as.Date("2024-12-31"))
humm_test <- humm_monthly %>% filter(date >= as.Date("2025-01-01"))

ts_humm_train <- ts(humm_train$mean_obs, start = c(2015, 1), frequency = 12)
humm_test_obs <- humm_test$mean_obs
```

After initial plotting, outlier removal was performed using tsclean() (Figure X).

```{r humm_clean, echo=FALSE, fig.width=12, fig.height=5, fig.cap="Initial time series and outlier removal for monthly Ruby-throated Hummingbird data"}
plot1 <- autoplot(ts_humm_train) +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Monthly Hummingbird Data") +
  theme_bw(base_size = 13) +
  scale_x_continuous(breaks = seq(2015, 2025, by = 1))

ts_humm_cleaned <- tsclean(ts_humm_train)

plot2 <- autoplot(ts_humm_train, series = "Original Data") +
  autolayer(ts_humm_cleaned, series = "Cleaned Data") +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Outlier Removal: Monthly Hummingbird Data") +
  theme_bw(base_size = 13) +
  theme(legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.title = element_blank())

plot_grid(plot1, plot2, ncol = 2)
```

After cleaning, summary statistics were calculated. The data shows strong seasonal variation with observations concentrated in summer months and near-zero in winter when hummingbirds migrate south.

```{r humm_stats, echo=FALSE}
stats_table <- data.frame(mean_obs = as.numeric(ts_humm_cleaned)) %>%
  summarise(
    Mean = mean(mean_obs, na.rm = TRUE),
    Median = median(mean_obs, na.rm = TRUE),
    SD = sd(mean_obs, na.rm = TRUE),
    Min = min(mean_obs, na.rm = TRUE),
    Max = max(mean_obs, na.rm = TRUE),
    `1st Quartile` = quantile(mean_obs, 0.25, na.rm = TRUE),
    `3rd Quartile` = quantile(mean_obs, 0.75, na.rm = TRUE)
  )

kbl(stats_table,
    caption = "Summary Statistics for Cleaned Monthly Hummingbird Observations",
    digits = 2,
    booktabs = TRUE) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  )
```

To visualize the seasonal pattern, I plotted monthly distributions and decomposed the series.

```{r humm_explore, echo=FALSE, fig.height=4, fig.width=14, fig.cap="Monthly distributions and decomposition of Ruby-throated Hummingbird time series"}
humm_train <- humm_train %>%
  mutate(month = factor(month(date), levels = 1:12))

boxplot <- ggplot(humm_train, aes(x = month, y = mean_obs)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  xlab("Month") +
  ylab("Mean Daily Observations") +
  ggtitle("Monthly Hummingbird Observations (2015-2024)") +
  scale_x_discrete(labels = month.abb) +
  theme_bw(base_size = 13)

humm_decomp <- decompose(ts_humm_cleaned, type = "additive")
decomp_plot <- autoplot(humm_decomp) +
  ggtitle("Decomposition of Cleaned Hummingbird Data") +
  theme_bw(base_size = 13)

plot_grid(boxplot, decomp_plot, ncol = 2)
```

The boxplot confirms the summer migrant status with observations peaking May-September and basically zero in winter. The decomposition shows a strong seasonal component with consistent annual pattern and no obvious long-term trend.

#### ACF and PACF

The ACF and PACF were plotted for initial modeling considerations.

```{r humm_acf, echo=FALSE, fig.cap="ACF and PACF for the cleaned monthly Ruby-throated Hummingbird series"}
acf_plot <- autoplot(Acf(ts_humm_cleaned, lag.max = 50, plot = FALSE)) +
  ggtitle(" ") +
  theme_bw(base_size = 13)

pacf_plot <- autoplot(Pacf(ts_humm_cleaned, lag.max = 50, plot = FALSE)) +
  ggtitle(" ") +
  theme_bw(base_size = 13)

plot_row <- plot_grid(acf_plot, pacf_plot, ncol = 2)

title <- ggdraw() + 
  draw_label("Cleaned Hummingbird Data: ACF and PACF",
    x = 0, hjust = 0, vjust = 1) +
  theme(plot.margin = margin(0, 0, 0, 0))

plot_grid(title, plot_row, ncol = 1, rel_heights = c(0.09, 1))
```

The ACF shows strong seasonal autocorrelation with significant spikes at lags 12, 24, and 36. The gradual decay suggests the series is non-stationary. The PACF has significant spikes at lags 1 and 12.

```{r humm_diff_check, echo=FALSE}
cat("ndiffs:", ndiffs(ts_humm_cleaned), "\n")
cat("nsdiffs:", nsdiffs(ts_humm_cleaned), "\n")
```

The functions suggested d=0 and D=1. Based on the strong seasonality, I used SARIMA models.

#### Model Fitting

I fit several SARIMA models manually based on the ACF/PACF, then compared to auto.arima().

```{r humm_models, echo=FALSE}
# Manual models based on ACF/PACF interpretation
Model_101_110 <- Arima(ts_humm_cleaned, order=c(1,0,1), seasonal=c(1,1,0))
Model_001_110 <- Arima(ts_humm_cleaned, order=c(0,0,1), seasonal=c(1,1,0))
Model_100_110 <- Arima(ts_humm_cleaned, order=c(1,0,0), seasonal=c(1,1,0))

# Auto
Model_auto <- auto.arima(ts_humm_cleaned, seasonal = TRUE)

compare_aic <- data.frame(
  Model = c("ARIMA(1,0,1)(1,1,0)[12]", 
            "ARIMA(0,0,1)(1,1,0)[12]",
            "ARIMA(1,0,0)(1,1,0)[12]",
            "Auto-SARIMA"),
  AIC = c(Model_101_110$aic, Model_001_110$aic, 
          Model_100_110$aic, Model_auto$aic)
)

kbl(compare_aic,
    caption = "AIC Comparison for SARIMA Models",
    digits = 2,
    booktabs = TRUE,
    row.names = FALSE) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center",
    stripe_index = which.min(compare_aic$AIC)
  )
```

Auto.arima selected ARIMA(2,0,1)(1,1,0)[12] with the lowest AIC of 366.53.

```{r humm_residuals, include=FALSE}
checkresiduals(Model_auto)
```

```{r humm_resid_plot, echo=FALSE, fig.cap="Residual diagnostics for selected SARIMA model"}
plot_grid(
  autoplot(Model_auto$residuals) + theme_bw(base_size = 11),
  autoplot(Acf(Model_auto$residuals, lag.max = 40, plot = FALSE)) + theme_bw(base_size = 11),
  autoplot(Pacf(Model_auto$residuals, lag.max = 40, plot = FALSE)) + theme_bw(base_size = 11),
  nrow = 1
)
```

The residuals look like white noise and the Ljung-Box test p-value was 0.39, so the model passes diagnostics.

#### Forecasting

I forecast 2025 using SARIMA and seasonal naive as a benchmark.

```{r humm_forecast, echo=FALSE}
sarima_forecast <- forecast(Model_auto, h = 12)
naive_forecast <- snaive(ts_humm_cleaned, h = 12)
```

```{r humm_compare, echo=FALSE, fig.cap="Forecast comparison for 2025 Ruby-throated Hummingbird observations"}
ts_humm_test <- ts(humm_test_obs, start = c(2025, 1), frequency = 12)

autoplot(ts_humm_test, series = "True Observations") +
  autolayer(sarima_forecast, PI = FALSE, series = "Auto-SARIMA") +
  autolayer(naive_forecast, PI = FALSE, series = "Seasonal Naive") +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Forecast Comparison: Test Period (2025)") +
  theme_bw(base_size = 13) +
  xlim(2024, 2026) +
  scale_x_continuous(breaks = seq(2024, 2026, by = 1)) +
  theme(legend.position = c(0.15, 0.88),
        legend.title = element_blank())
```

Both models capture the seasonal pattern. To check accuracy:

```{r humm_accuracy, echo=FALSE}
sarima_scores <- accuracy(sarima_forecast$mean, humm_test_obs)
naive_scores <- accuracy(naive_forecast$mean, humm_test_obs)

scores <- as.data.frame(rbind(sarima_scores, naive_scores))
row.names(scores) <- c("Auto-SARIMA", "Seasonal Naive")

kbl(scores[, c("ME", "RMSE", "MAE", "MPE", "MAPE")],
    caption = "Forecast Accuracy for Monthly Ruby-throated Hummingbird Observations",
    digits = 2,
    booktabs = TRUE) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center",
    stripe_index = which.min(scores[,"RMSE"])
  )

```

Seasonal naive achieved perfect accuracy with RMSE of 0.00 compared to 0.15 
for SARIMA. The seasonal naive method outperformed SARIMA on the test set 
because 2025 hummingbird observations followed an almost identical pattern to 
2024. This makes sense for a migratory species with consistent timing: the 
birds arrive and depart on similar schedules each year, which is exactly what 
seasonal naive assumes. MPE and MAPE show NaN/Inf for seasonal naive and 
-Inf/Inf for SARIMA because of division by zero when actual observations are 
zero in winter months.

A known limitation of citizen science data is that observation counts reflect 
both bird abundance and observer effort. Our above analysis of 
Yellow-rumped Warbler controlled for this by modeling observer ratio 
(proportion of observers detecting the species) and found similar seasonal 
patterns in both metrics, suggesting the trends are ecologically meaningful 
rather than artifacts of sampling effort. The consistent seasonal pattern 
observed in hummingbird data likely reflects true migration timing rather 
than observer bias.


# Summary and Conclusions

This project analyzed time series data from three bird species observed in Durham, NC — the Canada Goose, the Yellow-rumped Warbler (YRWA), and the Ruby-throated Hummingbird — spanning 2015 to 2025. For each species, we explored seasonal structure, fit multiple forecasting models, and evaluated out-of-sample performance. Several consistent findings emerged across all three analyses.

All three species exhibited strong and consistent seasonal patterns, though the nature and timing of those patterns differed in ecologically meaningful ways. The Canada Goose, as a resident species, showed year-round observations with modest seasonal fluctuation. The Yellow-rumped Warbler, a winter migrant, displayed sharp peaks during its wintering months and near-zero counts in summer. The Ruby-throated Hummingbird showed the most pronounced seasonality of the three, with observations concentrated entirely in the summer breeding season (May–September) and essentially zero presence in winter months. These patterns were confirmed by decomposition, ACF/PACF plots, and seasonal boxplot distributions across all three species.

Across all three species, SARIMA models consistently performed best or among the best in out-of-sample forecasting. The model's ability to account for both short-term autoregressive dependence and seasonal structure made it well-suited to these data. For the Canada Goose, Auto-SARIMA(1,1,1)(1,0,0)[12] outperformed the Basic Structural Model across RMSE, MAE, and MAPE metrics. For the YRWA, SARIMA achieved the lowest RMSE (12.60), MAE (9.81), and MASE (1.55) compared to ARIMA + Fourier and ETS models. For the Ruby-throated Hummingbird, while the seasonal naive benchmark matched performance on the 2025 test set due to the species' highly regular annual timing, the SARIMA model ARIMA(2,0,1)(1,1,0)[12] passed all residual diagnostics and produced a sound 12-month forward forecast.

## Model Performance

Different models worked best for each species:

**Canada Goose:** Auto-SARIMA(1,1,1)(1,0,0)[12] outperformed the Basic Structural Model across all error metrics, with an RMSE of 26.10 compared to 29.07 for the BSM. Both models captured the general seasonal shape but underestimated the magnitude of the seasonal dip in early 2025.

**Yellow-rumped Warbler:** SARIMA ARIMA(1,0,0)(2,1,0)[12] performed best with RMSE 12.60, MAE 9.81, and MASE 1.55, outperforming both the ARIMA + Fourier and ETS models. All three models captured the timing of seasonal peaks and troughs but differed in their accuracy of magnitude.

**Ruby-throated Hummingbird:** Seasonal naive got perfect accuracy (RMSE = 0.00) compared to SARIMA (RMSE = 0.15). The 2025 pattern was almost identical to 2024, which makes sense for a migrant species with consistent timing. SARIMA slightly overpredicted the peak, probably because it's averaging across all 10 years of training data.

## Observer Effort

One issue with citizen science data is that more observations could mean more birds or just more people birding. The warbler analysis checked this by comparing raw counts to observer ratio (how many observers saw the species). Both showed the same seasonal patterns, which suggests we're seeing real bird patterns and not just sampling bias.

We only analyzed observation counts for goose and hummingbird, but the consistent patterns across species and the warbler validation suggest the trends are real.

## Main Takeaways

All three species showed clear seasonal patterns that matched their life histories - year-round with variation (goose), summer only (hummingbird), winter only (warbler). Simple models worked well when patterns were stable year-to-year, while SARIMA was better for more variable species. The strong seasonal signals show that long-term eBird data is useful for tracking bird phenology despite the observer effort limitations.

Overall, this project demonstrates that time series methods, particularly SARIMA, provide a practical and interpretable framework for modeling avian observation data. The strong seasonal patterns present in all three species make them well-suited to seasonal time series approaches, and the forecasting results suggest these methods can meaningfully capture the temporal dynamics of bird populations in the Durham region.


