---
title: "Ruby-throated Hummingbird Time Series Analysis"
author: "Jonathan Gilman"
date: "2026-04-06"
output:
  pdf_document:
    toc: true
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)
library(tidyverse)
library(cowplot)
library(kableExtra)
```

## Introduction, Motivation, Relevance, and Objectives

Write a few paragraphs detailing the rationale for your study...

## Dataset Information

Provide information on how the dataset for this analysis were collected...

```{r}
#read data
humm_monthly <- read.csv("./Data/Cleaned_bySpecies/Hummingbird_Monthly.csv")

#inspect data
head(humm_monthly)
nobs <- nrow(humm_monthly)

#format date column
humm_monthly <- humm_monthly %>%
  mutate(date = as.Date(date))
```

```{r}
#keep last 12 months for forecast accuracy checking
n_for <- 12

#filter training data: 2015-2024
humm_train <- humm_monthly %>%
  filter(date >= as.Date("2015-01-01") &
         date <= as.Date("2024-12-01"))

#filter test data: 2025
humm_test <- humm_monthly %>%
  filter(date >= as.Date("2025-01-01") &
         date <= as.Date("2025-12-01"))
```


```{r}
kbl(head(humm_train[, c("date","mean_obs")], 10),
    caption = "Monthly Ruby-throated Hummingbird observation data, 2015-2024.",
    col.names = c("Date", "Mean Daily Observations"),
    digits = 2) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = "hold_position")
```

## Analysis

### Transforming data into time series object

```{r}
#create training ts object
ts_humm_train <- ts(humm_train$mean_obs,
                    start = c(2015, 1),
                    frequency = 12)

#store test observations for accuracy()
humm_test_obs <- humm_test$mean_obs
```

### Initial Plots

```{r}
autoplot(ts_humm_train) +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Ruby-throated Hummingbird Observations (2015-2024)") +
  theme_bw(base_size = 13) +
  scale_x_continuous(breaks = seq(2015, 2025, by = 1))
```

```{r}
humm_train <- humm_train %>%
  mutate(month = factor(month(date), levels = 1:12))

ggplot(humm_train, aes(x = month, y = mean_obs)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  xlab("Month") +
  ylab("Mean Daily Observations") +
  ggtitle("Monthly Hummingbird Observations by Month (2015-2024)") +
  scale_x_discrete(labels = month.abb) +
  theme_bw(base_size = 13)
```

```{r}
#ACF and PACF of original training series
par(mfrow = c(1,2))
Acf(ts_humm_train, lag.max = 40, main = "Hummingbird ACF and PACF")
Pacf(ts_humm_train, lag.max = 40, main = "")
par(mfrow = c(1,1))
```

### Decomposing the time series

```{r}
humm_decomp <- decompose(ts_humm_train, type = "additive")
autoplot(humm_decomp) +
  ggtitle("Decomposition of Monthly Hummingbird Observations") +
  theme_bw(base_size = 13)
```

```{r}
#remove seasonal component
humm_deseasonal <- seasadj(humm_decomp)

#plot original vs deseasoned
autoplot(ts_humm_train, series = "Original") +
  autolayer(humm_deseasonal, series = "Deseasoned") +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Original and Deseasoned Hummingbird Data") +
  theme_bw(base_size = 13) +
  theme(legend.title = element_blank(),
        legend.position = c(0.15, 0.88))
```

```{r}
#compare ACFs: original vs deseasoned
plot_grid(
  autoplot(Acf(ts_humm_train, lag.max = 40, plot = FALSE),
           main = "ACF: Original") +
    theme_bw(base_size = 11),
  autoplot(Acf(humm_deseasonal, lag.max = 40, plot = FALSE),
           main = "ACF: Deseasoned") +
    theme_bw(base_size = 11),
  nrow = 1
)
```

```{r}
#compare PACFs: original vs deseasoned
plot_grid(
  autoplot(Pacf(ts_humm_train, lag.max = 40, plot = FALSE),
           main = "PACF: Original") +
    theme_bw(base_size = 11),
  autoplot(Pacf(humm_deseasonal, lag.max = 40, plot = FALSE),
           main = "PACF: Deseasoned") +
    theme_bw(base_size = 11),
  nrow = 1
)
```

### Stationarity Test

```{r}
#run ADF on deseasoned series
print(adf.test(humm_deseasonal, alternative = "stationary"))

#how many differences needed
n_diff  <- ndiffs(humm_deseasonal)
ns_diff <- nsdiffs(ts_humm_train)

cat("Non-seasonal differences needed:", n_diff, "\n")
cat("Seasonal differences needed:", ns_diff, "\n")
```

```{r}
#difference the deseasoned series
humm_deseasonal_diff <- diff(humm_deseasonal, differences = 1, lag = 1)

#plot original, deseasoned, and differenced
autoplot(ts_humm_train, series = "Original") +
  autolayer(humm_deseasonal, series = "Deseasoned") +
  autolayer(humm_deseasonal_diff, series = "Deseasoned + Diff") +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Original, Deseasoned, and Differenced Hummingbird Data") +
  theme_bw(base_size = 13) +
  theme(legend.title = element_blank(),
        legend.position = c(0.15, 0.88))
```

```{r}
#compare ACFs after differencing
plot_grid(
  autoplot(Acf(ts_humm_train, lag.max = 40, plot = FALSE),
           main = "ACF: Original") +
    theme_bw(base_size = 11),
  autoplot(Acf(humm_deseasonal, lag.max = 40, plot = FALSE),
           main = "ACF: Deseasoned") +
    theme_bw(base_size = 11),
  autoplot(Acf(humm_deseasonal_diff, lag.max = 40, plot = FALSE),
           main = "ACF: Deseasoned + Diff") +
    theme_bw(base_size = 11),
  nrow = 1
)
```

```{r}
#compare PACFs after differencing
plot_grid(
  autoplot(Pacf(ts_humm_train, lag.max = 40, plot = FALSE),
           main = "PACF: Original") +
    theme_bw(base_size = 11),
  autoplot(Pacf(humm_deseasonal, lag.max = 40, plot = FALSE),
           main = "PACF: Deseasoned") +
    theme_bw(base_size = 11),
  autoplot(Pacf(humm_deseasonal_diff, lag.max = 40, plot = FALSE),
           main = "PACF: Deseasoned + Diff") +
    theme_bw(base_size = 11),
  nrow = 1
)
```

### Fitting Models

#### Model 1: SARIMA (manual)

```{r}
##fit candidate SARIMA models and compare AIC
#orders informed by ACF/PACF of differenced deseasoned series

sarima_fit1 <- Arima(ts_humm_train, order = c(1,1,1), seasonal = c(1,1,0))
sarima_fit2 <- Arima(ts_humm_train, order = c(0,1,1), seasonal = c(0,1,1))
sarima_fit3 <- Arima(ts_humm_train, order = c(1,1,0), seasonal = c(0,1,1))
sarima_fit4 <- Arima(ts_humm_train, order = c(1,1,1), seasonal = c(1,0,0))

compare_aic <- data.frame(
  Model = c("SARIMA(1,1,1)(1,1,0)[12]",
            "SARIMA(0,1,1)(0,1,1)[12]",
            "SARIMA(1,1,0)(0,1,1)[12]",
            "SARIMA(1,1,1)(1,0,0)[12]"),
  AIC = c(sarima_fit1$aic,
          sarima_fit2$aic,
          sarima_fit3$aic,
          sarima_fit4$aic)
)
print(compare_aic)
```

```{r}
#check residuals for best manual model by AIC
best_manual_idx <- which.min(compare_aic$AIC)
sarima_manual_best <- list(sarima_fit1, sarima_fit2, sarima_fit3, sarima_fit4)[[best_manual_idx]]
cat("Best manual model:", compare_aic$Model[best_manual_idx], "\n")
checkresiduals(sarima_manual_best)
```



#### Model 2: Auto SARIMA

```{r}
#auto ARIMA on training series
sarima_auto_fit <- auto.arima(ts_humm_train)
print(sarima_auto_fit)
checkresiduals(sarima_auto_fit)
```

#### Model 3: Seasonal Naive

```{r}
#seasonal naive as benchmark
snaive_fit <- snaive(ts_humm_train, h = n_for)
checkresiduals(snaive_fit)
```

### Forecasting

```{r}
#generate 12-step-ahead forecasts
sarima_manual_for <- forecast(sarima_manual_best, h = n_for)
sarima_auto_for   <- forecast(sarima_auto_fit, h = n_for)

```

### Checking Accuracy

```{r}
#compute accuracy against test observations
sarima_manual_scores <- accuracy(sarima_manual_for$mean, humm_test_obs)
sarima_auto_scores   <- accuracy(sarima_auto_for$mean,   humm_test_obs)
snaive_scores        <- accuracy(snaive_fit$mean,        humm_test_obs)

#create comparison data frame
seas_scores <- as.data.frame(rbind(sarima_manual_scores,
                                    sarima_auto_scores,
                                    snaive_scores))
row.names(seas_scores) <- c(compare_aic$Model[best_manual_idx],
                             "ARIMA(1,0,0)(1,1,2)[12]",
                             "Seasonal Naive")

#choose model with lowest RMSE
best_model_index <- which.min(seas_scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(seas_scores[best_model_index,]))
```

```{r}
kbl(seas_scores[, c("ME", "RMSE", "MAE")],
    caption = "Forecast accuracy for monthly Ruby-throated Hummingbird observations",
    digits = array(2, 3)) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = "hold_position") %>%
  kable_styling(latex_options = "striped",
                stripe_index = which.min(seas_scores[,"RMSE"]))
```

```{r}
#plot all forecasts and observed values together
ts_humm_test <- ts(humm_test_obs, start = c(2025, 1), frequency = 12)

autoplot(ts_humm_test, series = "True Observations") +
  autolayer(sarima_manual_for, PI = FALSE,
            series = compare_aic$Model[best_manual_idx]) +
  autolayer(sarima_auto_for, PI = FALSE,
            series = "ARIMA(1,0,0)(1,1,2)[12]") +
  autolayer(snaive_fit, PI = FALSE,
            series = "Seasonal Naive") +
  xlab("Time") +
  ylab("Mean Daily Observations") +
  ggtitle("Forecast Comparison: Test Period (2025)") +
  theme_bw(base_size = 13) +
  guides(colour = guide_legend(title = "Model"))
```

